---
title: "Grading Student Project Data"
subtitle: "EDA and Data Preprocessing"
output:
  html_document:
    code_folding: show
    df_print: paged
    number_sections: yes
    theme: cerulean
    toc: yes
    toc_float: true
    code_download: true

---

```{r setup, include=FALSE,message=FALSE}
knitr::opts_chunk$set(echo = TRUE, message=FALSE,warning=FALSE)
knitr::opts_chunk$set(fig.width=6, fig.height=4)

if(require(pacman)==0)
   {install.packages("pacman")}
pacman::p_load(devtools,caret,fastDummies,tidyverse,skimr, GGally,DataExplorer,ggrepel,ggthemes,dslabs,rpart,rpart.plot,randomForest,xgboost,ROCR,knitr,rmarkdown)
```

# Load Data

```{r}
file_to_grade <- "df_flags_2nd_run.rds"
dat <- readRDS(file_to_grade)

cat("Loaded:", basename(file_to_grade), "\n")
dim(dat)
```

100 variables??? Yikes!

# Row Count Check

```{r}
original_n <- 237730  
student_n <- nrow(dat)
cat("Original rows:", original_n, "\n")
cat("Student rows:", student_n, "\n")
cat("Rows lost:", original_n - student_n, "\n")
```

You lost a nominal amount of data.


# Factor Check

All strings should be factors and the number of levels should be reasonable.

```{r}
dat %>%
  summarise(across(everything(), ~class(.)[1])) %>%
  pivot_longer(everything(), names_to = "Variable", values_to = "Type") %>%
  arrange(Type) %>% 
  print(n=Inf)

# Quick look at factor variables and their number of levels
dat %>%
  select(where(is.factor)) %>%
  summarise(across(everything(), ~nlevels(.))) %>%
  pivot_longer(everything(), names_to = "Variable", values_to = "Levels") %>%
  arrange(desc(Levels))
```

You need to transform or remove date before modeling.  Looks like everything is binned and converted to binary.  Interesting approach.  We will see how it works. No explanations in your RMD, so I can't really follow what you did.


# Check for missing

```{r}
plot_missing(dat)
```
No missing.

# Check Target

Should be a factor coded "No","Yes"

```{r}
class(dat$loan_default)
table(dat$loan_default)
levels(dat$loan_default)
```
Good!

# Check Correlations

Should not have any variables that are very highly correlated.

```{r}
dat %>% 
  select(where(is.numeric)) %>% 
  select(where(~!is.integer(.))) %>% 
  plot_correlation()
```

Hard to read the plot, but you have some very highly correlated variables in the data. If there are redundancies, some models will not run.

# Check Numerical Summaries

Checking for zero standard deviations, outliers.

```{r}
dat %>%
  select(where(is.numeric)) %>%
  #select(where(~!is.integer(.))) %>% 
  summarise(across(
    everything(),
    list(
      mean = ~mean(., na.rm = TRUE),
      sd = ~sd(., na.rm = TRUE),
      min = ~min(., na.rm = TRUE),
      q25 = ~quantile(., 0.25, na.rm = TRUE),
      median = ~median(., na.rm = TRUE),
      q75 = ~quantile(., 0.75, na.rm = TRUE),
      max = ~max(., na.rm = TRUE)
    ),
    .names = "{.col}__{.fn}"
  )) %>%
  pivot_longer(
    everything(),
    names_to = c("Variable", "Statistic"),
    names_sep = "__",
    values_to = "Value"
  ) %>%
  pivot_wider(names_from = Statistic, values_from = Value) %>%
  arrange(Variable) %>%
  kable(
    digits = 2,
    caption = "Numeric Variable Summary (Mean, SD, Min, 25th, Median, 75th, Max)"
  )
```

Some of your dummy variables are mostly 0 as indicated by the very small mean and std. dev.  YOu may have some challenges.

# Check Barplots

Should not have any categorical variables with only one (or nearly one) level

```{r}
plot_bar(dat,nrow=3,ncol=2)
```

Some very sparse levels.  This may mean some predictors won't be meaningful.  

# Check Histograms

Should not have any odd patterns or variables with only one (or very few) value(s).

```{r}
plot_histogram(dat,nrow=3,ncol=2)
```

There are some problems here with sparsity. You have transformed a lot of variables.  This may be ok.  But if your predictions are not good, consider going back to the raw data.


# Final comments

There are still a few problems with your data that will make modeling a challenge.  I recommend taking another pass at cleaning up the data before you start.  I tried to follow along with your markdown.  While it is organized, there is limited logic or annotation noted,and I had difficulty understanding your reasoning.  