---
title: "group11a_Chlebowy_Shane_Trees"
output: html_document
date: "2025-11-06"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


# Reading the data
Prior to this, we applied our data cleaning on the holdout sample and addressed the issue with an ANY level appearing in the home_ownership variable
```{r}
file <- "train_trees.rds"
df_trees <- readRDS(file)

holdout = "holdout_trees.rds"
holdout_trees = readRDS(holdout)

# Load all packages used in this analysis
library(rpart)
library(rpart.plot)
library(caret)
library(MLmetrics)
library(ROCR)
library(dplyr)
library(ggplot2)  # optional but loaded by caret/ROCR plots


```


# Undersampling the data
Balancing Yes/No's in the training data to help build a more generalizable models that can better predict on unseen data. We fit some trees with the balanced data, and some with the raw data 
```{r}

set.seed(123)

df_trees_balanced = df_trees %>%
  group_by(loan_default) %>%
  sample_n(size = min(table(df_trees$loan_default))) %>%
  ungroup()

table(df_trees_balanced$loan_default) # it is balanced

saveRDS(df_trees_balanced, "df_trees_balanced.rds")

```


# TREE 1 - Basic Tree

```{r}
set.seed(123)

TREE_1 <- rpart::rpart(
  loan_default ~ .,
  data = df_trees,
  method = "class",
  control = rpart::rpart.control(cp = 0, minsplit = 5000, minbucket = 2500) 
)

options(scipen = 999)

rpart.plot::rpart.plot(TREE_1, extra = 1, fallen.leaves = TRUE) ## Very basic

```
Almost all of the splits are on dates. I believe this could be a result of too strict controls

## Predictions and Confusion Matrix
```{r}

Tree1_train_pred = predict(TREE_1,
                    df_trees,
                    type = "class")


tree_preds = data.frame( Actuals = holdout_trees$loan_default)

tree_preds$Tree1_class = predict(TREE_1,
                              holdout_trees,
                              type = "class")
tree_preds$Tree1_prob = predict(TREE_1,
                              holdout_trees,
                              type = "prob")[,"Yes"]

confusionMatrix(Tree1_train_pred,
                df_trees$loan_default,
                positive = "Yes")

Tree1_hold_pred = predict(TREE_1,
                    holdout_trees,
                    type = "class")
confusionMatrix(Tree1_hold_pred,
                holdout_trees$loan_default,
                positive = "Yes")

```
Very ineffective model. Low recall (.33) and only a 2% improvement to the No Information Rate. Not a very good tree at all


# TREE 2 -- Cross-validated Tree


## Cross Validating
cross-validating to give the model more train/test splits and to optimize cp. It should make for a more accurate and more generalizable model. Adjusted controls to fit deeper trees
```{r} 
set.seed(123)

CV_TREE <- rpart::rpart(
  loan_default ~ .,
  data = df_trees,
  method = "class",
  control = rpart::rpart.control(cp = 0, xval = 10, minsplit = 500, minbucket = 250, maxdepth = 25)
) ## changed controls to fit deeper trees 

printcp(CV_TREE)
plotcp(CV_TREE)


best_cp <- CV_TREE$cptable[which.min(CV_TREE$cptable[,"xerror"]), "CP"]
print(best_cp) ## 0.002198061


```


## Pruning TREE2
```{r}
TREE_2 <- prune(CV_TREE, cp = best_cp)

options(scipen = 999)

rpart.plot::rpart.plot(TREE_2, extra = 1, fallen.leaves = TRUE)
```
This Resulted in a much deeper tree, and it looks like it is splitting on more than just the date variables.

## Predictions and ConfusionMatrix
```{r}

Tree2_train_pred = predict(TREE_2,
                    df_trees,
                    type = "class")


tree_preds$Tree2_class = predict(TREE_2,
                              holdout_trees,
                              type = "class")
tree_preds$Tree2_prob = predict(TREE_2,
                             holdout_trees,
                             type = "prob")[,"Yes"]



confusionMatrix(Tree2_train_pred,
                df_trees$loan_default,
                positive = "Yes")

Tree2_hold_pred = predict(TREE_2,
                    holdout_trees,
                    type = "class")


confusionMatrix(Tree2_hold_pred,
                holdout_trees$loan_default,
                positive = "Yes")

```
Improved overall accuracy and detection rate slightly, but still only a 3% gain from the No Information Rate. Recall is still only 38%. Not a very good model either


# TREE 3 - Cross-Validated Tree with Balanced Data
We changed controls again, allowing for smaller nodes
## Cross Validating
```{r}

set.seed(123)

CV2_TREE <- rpart::rpart(
  loan_default ~ .,
  data = df_trees_balanced,
  method = "class",
  control = rpart::rpart.control(cp = .0002, xval = 10, minsplit = 200, minbucket = 100, maxdepth = 15)
)

printcp(CV2_TREE)
plotcp(CV2_TREE)


best_cp2 <- CV2_TREE$cptable[which.min(CV2_TREE$cptable[,"xerror"]), "CP"]
print(best_cp2) #0.000214445
```

## Pruning TREE 3
```{r}


TREE_3 <- prune(CV2_TREE, cp = best_cp2)

options(scipen = 999)

rpart.plot::rpart.plot(TREE_3, extra = 1, fallen.leaves = TRUE)

```
Very deep tree, appears to have more depth than the previous

## Predictions and ConfusionMatrix

```{r}

Tree3_train_pred = predict(TREE_3,
                    df_trees_balanced,
                    type = "class")

Tree3_hold_pred = predict(TREE_3,
                    holdout_trees,
                    type = "class")


tree_preds$Tree3_class = predict(TREE_3,
                              holdout_trees,
                              type = "class")
tree_preds$Tree3_prob = predict(TREE_3,
                             holdout_trees,
                             type = "prob")[,"Yes"]

confusionMatrix(Tree3_train_pred,
                df_trees_balanced$loan_default,
                positive = "Yes") 

confusionMatrix(Tree3_hold_pred,
                holdout_trees$loan_default,
                positive = "Yes") 
```
Significantly better model. Overall Accuracy is 76%, a 26% jump from a no information rate of 50%. Recall of .7745 is significantly better compared to previous models. Only able to correctly ID 38% of Yes's, but much better than previous models.



# TREE 4 -- Balanced cv Tree using "information"
Experimenting using information instead of Gini to split the trees

```{r}

set.seed(123)

CV3_TREE <- rpart::rpart(
  loan_default ~ .,
  data = df_trees_balanced,
  method = "class",
  parms = list(split = "information"),
  control = rpart::rpart.control(
    cp = 0.0002,
    xval = 10,
    minsplit = 200,
    minbucket = 100,
    maxdepth = 15
  )
)


printcp(CV3_TREE)
plotcp(CV3_TREE)


best_cp3 <- CV3_TREE$cptable[which.min(CV3_TREE$cptable[,"xerror"]), "CP"]
print(best_cp3) #0.000343112

```

## Pruning Tree 4

```{r}

TREE_4 <- prune(CV3_TREE, cp = best_cp3)
options(scipen = 999)
rpart.plot::rpart.plot(TREE_4, extra = 1, fallen.leaves = FALSE)

```
A shallower tree then other previous trees

## Predictions and Confusion Matrix
```{r}

Tree4_train_pred = predict(TREE_4,
                    df_trees_balanced,
                    type = "class")
Tree4_hold_pred = predict(TREE_4,
                    holdout_trees,
                    type = "class")

tree_preds$Tree4_class = predict(TREE_4,
                               holdout_trees,
                               type = "class")
tree_preds$Tree4_prob = predict(TREE_4,
                              holdout_trees,
                              type = "prob")[,"Yes"]

confusionMatrix(Tree4_train_pred,
                df_trees_balanced$loan_default,
                positive = "Yes") 

confusionMatrix(Tree4_hold_pred,
                holdout_trees$loan_default,
                positive = "Yes")


```

Very similar model compared to the last. Overall Accuracy is 75.5%, and recall of 76%. Detection rate took a small dip. Less accuracy and sensitivity drop on the holdout, so potentially a slightly better model than Tree 3

# TREE 5 -- Using Caret, optimizing threshold
trying some new things, like using a tune grid to find the best cp, and using reapeatcv to get even more cross validation. We also optimize threshold for this tree trying to improve recall

```{r}
set.seed(123)

grid <- expand.grid(
  cp = seq(0.0002, 0.002, by = 0.00005)  # Range of cp's to try
)

train_control <- trainControl(
  method = "repeatedcv", number = 10, repeats = 3,  # fitting 10 models at each split, 3 times (30 trees for each cp split) should help us find the best possible model
  summaryFunction = twoClassSummary,
  classProbs = TRUE,
  savePredictions = TRUE
)

CV5_TREE <- train(
  loan_default ~ .,
  data = df_trees_balanced,
  method = "rpart",
  trControl = train_control,
  tuneGrid = grid,
  metric = "ROC"
)

print(CV5_TREE$bestTune)
plot(CV5_TREE)


best_cp <- CV5_TREE$bestTune$cp


```
## Fitting and Pruning Tree 5
```{r}

TREE_5 <- rpart(
  loan_default ~ .,
  data = df_trees_balanced,
  method = "class",
  parms = list(split = "information"),
  control = rpart.control(
    cp = best_cp,
    minsplit = 50,      # lower to allow finer splits
    minbucket = 25,     
    maxdepth = 20,       # a bit deeper than Tree 4
    xval = 10
  )
)
TREE_5_pruned <- prune(TREE_5, cp = best_cp)



```

## optimizing threshold for best F1 score
```{r}

Tree5_prob <- predict(TREE_5_pruned, df_trees_balanced, type = "prob")[,"Yes"]

# Optimize F1 threshold
thresholds <- seq(0.1, 0.9, by = 0.01)
f1_scores <- sapply(thresholds, function(t) {
  preds <- ifelse(Tree5_prob > t, "Yes", "No")
  MLmetrics::F1_Score(y_pred = preds, y_true = df_trees_balanced$loan_default, positive = "Yes")
})
best_thresh <- thresholds[which.max(f1_scores)]
best_f1 <- max(f1_scores)
cat(sprintf("Best F1 threshold: %.3f | Max F1: %.3f\n", best_thresh, best_f1))


Tree5_prob <- predict(TREE_5_pruned, holdout_trees, type = "prob")[,"Yes"]
Tree5_class <- ifelse(Tree5_prob > best_thresh, "Yes", "No")

tree_preds$Tree5_class = Tree5_class
tree_preds$Tree5_prob = Tree5_prob

Tree5_train_pred = predict(TREE_5_pruned,
                    df_trees_balanced,
                    type = "class")
Tree5_hold_pred = predict(TREE_5_pruned,
                    holdout_trees,
                    type = "class")
confusionMatrix(Tree5_train_pred,
                df_trees_balanced$loan_default,
                positive = "Yes") 
confusionMatrix(Tree5_hold_pred,
                holdout_trees$loan_default,
                positive = "Yes") 

```

Optimal threshold at .38

Opted to decrease controls even smaller to make even larger trees, trees with too few splits have been almost exclusively using dates (last pull, issue date). 76% overall accuracy, 77% recall, and still about 38% detection rate (.3846 -- highest yet). .748 Overall accuracy on holdout data (best yet)


# TREE 6 -- Optimizing Tree4 using caret + f1 optimized thresholds
going back to using the unbalanced data just to see if these better models work better with the unbalanced data set

```{r}

set.seed(123)
CV6_TREE <- rpart::rpart(
  loan_default ~ .,
  data = df_trees_balanced,
  method = "class",
  parms = list(split = "information"),   # use information gain
  control = rpart::rpart.control(
    cp = 0.0002,
    xval = 10,
    minsplit = 200,
    minbucket = 100,
    maxdepth = 10
  )
)

# Review complexity table and select optimal cp
printcp(CV6_TREE)
plotcp(CV6_TREE)

best_cp6 <- CV6_TREE$cptable[which.min(CV6_TREE$cptable[, "xerror"]), "CP"]
print(best_cp6)
```
## Plotting and Pruning
```{r}
TREE_6 <- prune(CV6_TREE, cp = best_cp6)
options(scipen = 999)
rpart.plot::rpart.plot(TREE_6, extra = 1, fallen.leaves = TRUE)
```
```{r}
# --- Step 3: Get predicted probabilities for holdout sample ---
Tree6_probs <- predict(TREE_6, holdout_trees, type = "prob")[, "Yes"]

# --- Step 4: Function to compute F1 across thresholds ---
find_best_f1 <- function(probs, actuals, positive = "Yes") {
  thresholds <- seq(0.1, 0.9, by = 0.01)
  
  f1s <- sapply(thresholds, function(th) {
    preds <- factor(ifelse(probs > th, "Yes", "No"), levels = c("No", "Yes"))
    cm <- confusionMatrix(preds, actuals, positive = positive)
    prec <- cm$byClass["Pos Pred Value"]
    rec <- cm$byClass["Sensitivity"]
    if (is.na(prec) | is.na(rec) | (prec + rec == 0)) return(0)
    2 * prec * rec / (prec + rec)
  })
  
  best_th <- thresholds[which.max(f1s)]
  best_f1 <- max(f1s)
  
  cat("Best Threshold:", round(best_th, 2), "| F1:", round(best_f1, 4), "\n")
  
  return(list(threshold = best_th, f1 = best_f1))
}
```

```{r}
best_T6 <- find_best_f1(Tree6_probs, holdout_trees$loan_default)
best_threshold <- best_T6$threshold

# --- Step 6: Apply optimal threshold ---
tree_preds$Tree6_class <- factor(ifelse(Tree6_probs > best_threshold, "Yes", "No"),
                               levels = c("No", "Yes"))
tree_preds$Tree6_prob  <- Tree6_probs


Tree6_train_pred = predict(TREE_6,
                    df_trees_balanced,
                    type = "class")
Tree6_hold_pred = predict(TREE_6,
                    holdout_trees,
                    type = "class")

confusionMatrix(Tree6_train_pred,
                df_trees_balanced$loan_default,
                positive = "Yes") 
confusionMatrix(Tree6_hold_pred,
                holdout_trees$loan_default,
                positive = "Yes") 

```
.64 chosen as the optimal threshold. Similar performance to other models 76% overall accuracy, .76 sensitivity, slightly lower detection rate. Best overall accuracy on holdout predictions, but also the lowest sensitivity


# ROC Curve
```{r}
# saving prediction data frame
saveRDS(tree_preds, "tree_preds.rds")

pred_Tree1 = ROCR::prediction(tree_preds$Tree1_prob,
                             tree_preds$Actuals)
roc_Tree1 = ROCR::performance(pred_Tree1,"tpr", "fpr")


pred_Tree2 = ROCR::prediction(tree_preds$Tree2_prob,
                             tree_preds$Actuals)
roc_Tree2 = ROCR::performance(pred_Tree2,"tpr", "fpr")


pred_Tree3 = ROCR::prediction(tree_preds$Tree3_prob,
                              tree_preds$Actuals)
roc_Tree3 = ROCR::performance(pred_Tree3,"tpr", "fpr")


pred_Tree4 = ROCR::prediction(tree_preds$Tree4_prob,
                              tree_preds$Actuals)
roc_Tree4 = ROCR::performance(pred_Tree4,"tpr", "fpr")

pred_Tree5 = ROCR::prediction(tree_preds$Tree5_prob,
                              tree_preds$Actuals)
roc_Tree5 = ROCR::performance(pred_Tree5,"tpr", "fpr")

pred_Tree6 = ROCR::prediction(tree_preds$Tree6_prob,
                              tree_preds$Actuals)
roc_Tree6 = ROCR::performance(pred_Tree6,"tpr", "fpr")

plot(roc_Tree1, col = "blue", lwd = 2 , main = "ROC Curves")
plot(roc_Tree2, col = "green", lwd = 2, add = TRUE)
plot(roc_Tree3, col = "orange", lwd = 2,  add = TRUE)
plot(roc_Tree4, col = "cyan", lwd = 2,  add = TRUE)
plot(roc_Tree5, col = "purple", lwd = 2,  add = TRUE)
plot(roc_Tree6, col = "yellow", lwd = 2,  add = TRUE)

abline(a = 0, b = 1, col = "red", lty = 2, lwd = 2)

```
Hard to say which is the best based off of ROC curve. Model 4,5,6 are all very similar 

## COMPARING AUC
```{r}

roc_Tree1_auc  <- ROCR::performance(pred_Tree1, 'auc')@y.values[[1]]
roc_Tree2_auc <- ROCR::performance(pred_Tree2, 'auc')@y.values[[1]]
roc_Tree3_auc <- ROCR::performance(pred_Tree3, 'auc')@y.values[[1]]
roc_Tree4_auc <- ROCR::performance(pred_Tree4, 'auc')@y.values[[1]]
roc_Tree5_auc <- ROCR::performance(pred_Tree5, 'auc')@y.values[[1]]
roc_Tree6_auc <- ROCR::performance(pred_Tree6, 'auc')@y.values[[1]]

cat("\nAUC Comparison:\n",
    sprintf("Tree 1: %.4f\n", roc_Tree1_auc),
    sprintf("Tree 2:    %.4f\n", roc_Tree2_auc),
    sprintf("Tree 3:    %.4f\n", roc_Tree3_auc),
    sprintf("Tree 4:    %.4f\n", roc_Tree4_auc),
    sprintf("Tree 5:    %.4f\n", roc_Tree5_auc),
    sprintf("Tree 6:    %.4f\n", roc_Tree6_auc)
)

```

Tree 6 Has the best AUC, but not my much. Tree 3,4,5,6 are all very close 


# Computing Lift 

```{r}

lift_Tree1 = ROCR::performance(pred_Tree1,"lift", "rpp")
lift_Tree2 = ROCR::performance(pred_Tree2,"lift", "rpp")
lift_Tree3 = ROCR::performance(pred_Tree3,"lift", "rpp")
lift_Tree4 = ROCR::performance(pred_Tree4,"lift", "rpp")
lift_Tree5 = ROCR::performance(pred_Tree5,"lift", "rpp")
lift_Tree6 = ROCR::performance(pred_Tree6,"lift", "rpp")


plot(lift_Tree1, main = "Lift Chart (unbalanced)", col="blue")
plot(lift_Tree2, add= TRUE, col="green")

plot(lift_Tree3, main = "Lift Chart (balanced)", col="orange")
plot(lift_Tree4, add= TRUE, col="cyan")
plot(lift_Tree5, add= TRUE, col="purple")
plot(lift_Tree6, add= TRUE, col="yellow")

```

They are all very similar. Tree 5 takes a sharp jump up to the other 3 models. I believe this might indicate that this model is predicting high probabilities for loans that actually did default, which would be a serious problem. Tree 5 might not be the best choice 

# Comparing F1
```{r}

f1_from_cm <- function(pred, actual) {
  cm <- confusionMatrix(factor(pred, levels=c("No","Yes")),
                        factor(actual, levels=c("No","Yes")),
                        positive="Yes")
  precision <- cm$byClass["Pos Pred Value"]
  recall <- cm$byClass["Sensitivity"]
  2 * ((precision * recall) / (precision + recall))
}

f1_Tree2 <- f1_from_cm(tree_preds$Tree2_class, tree_preds$Actuals)
f1_Tree3 <- f1_from_cm(tree_preds$Tree3_class, tree_preds$Actuals)
f1_Tree4 <- f1_from_cm(tree_preds$Tree4_class, tree_preds$Actuals)
f1_Tree5 <- f1_from_cm(tree_preds$Tree5_class, tree_preds$Actuals)
f1_Tree1  <- f1_from_cm(tree_preds$Tree1_class,  tree_preds$Actuals)
f1_Tree6  <- f1_from_cm(tree_preds$Tree6_class,  tree_preds$Actuals)



data.frame(Model=c("Tree1","Tree2","Tree3","Tree4","Tree5", "Tree6"),
           F1=c(f1_Tree1,f1_Tree2,f1_Tree3,f1_Tree4,f1_Tree5,f1_Tree6))

```

Tree 6 has the best F1 score. .55 is not a great F1. We will look to improve with Random Forest and boosted trees

While Tree 6 has the best F1, i am not convinced it is the best model. Tree 4 misclassified fewer "No"'s as "Yes"'s than Tree 6, had a better sensitivity than Tree 6, and a marginally different AUC compared to Tree 6.

## Saving The Best Tree
```{r}

saveRDS(TREE_4, "best_tree.rds")


```

