---
title: "Group11a_Chlebowy_Shane_NN"
output: html_document
date: "2025-12-02"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Loading Packages
```{r}
library(dplyr)
library(caret)
library(parallel)
library(doParallel)
library(pROC)
library(MLmetrics)
library(ROCR)
library(recipes)
library(ranger)

df_flags_2nd_run = readRDS("df_flags_2nd_run.rds")
holdout = readRDS("holdout.rds")

```


# Checking RF varimporatce to reduce variables
## Going to keep the top 15
```{r}
rf = readRDS('best_random_forest.rds')

varImp(rf)
```






# Preprocessing the data 
```{r}
# limiting to top 15 by RF varImp
df_flags_2nd_run = df_flags_2nd_run %>% select(loan_default, int_rate, issue_date_year, last_credit_pull_year, dti_cap38, loan_to_income_ratio_cap, revol_util_cap100, credit_age_months, revol_bal_cap100k, grade, fico_avg, loan_amnt, annual_inc_cap, total_acc_cap60, open_acc_cap25, emp_occupation)

# recipe
rec <- recipe(loan_default ~ ., data = df_flags_2nd_run) %>%
  step_range(all_numeric_predictors()) %>%
  step_dummy(all_nominal_predictors(), one_hot = FALSE)

prep_train <- prep(rec)
train_prepped <- bake(prep_train, new_data = NULL)
train_prepped
test_prepped <- bake(prep_train, new_data = holdout)


train_prepped = train_prepped %>% mutate(
  loan_default = factor(loan_default, levels = c("Yes","No"))
)

test_prepped = test_prepped %>% mutate(
  loan_default = factor(loan_default, levels = c("Yes","No"))
)

table(train_prepped$loan_default)
table(test_prepped$loan_default) # POSITIVE CLASSES ARE FIRST !!
```

```{r}
#cores = parallel::detectCores()
#cores
#cl = parallel::makeCluster(cores-1)
#registerDoParallel(cl)

#set.seed(123)

#tuneGrid = expand.grid(
#  size = seq(8,10, by = 1),
#  decay = seq(.3,.7, by = .1)    # Regularization strength
#)

#trControl = trainControl(
#  method = "cv",
#  number = 5,
#  classProbs = TRUE,
#  summaryFunction = twoClassSummary
#)
#nn=train(loan_default ~ .,
#                 data = train_prepped,
#                  method = "nnet",
#                  trControl= trControl,
#                  metric="ROC",
#                  linout = FALSE,
#                  trace = FALSE,
#                  tuneGrid = tuneGrid)
#nn

#saveRDS(nn, "nn_model1.rds")

#stopCluster(cl)
#registerDoSEQ()

```
After running a basic grid ( sizes 0-10, and decay 0-1), best tune was size 9, decay .4

2nd fit using grid (size 8,9,10 and Decay .3-.6) -- Model best params were size = 10 and decay = 0.7
I will test a wider set of params in my next test to narrow down where performance is best, since results varied after shrinking grid

- Sizes 5, 10, 15, 20
- Decay .1, .3, .5, .7 


## Making Predictions
### f1 optimized predictions

```{r}
nn = readRDS("nn_model1.rds")

# Creating Prediction Df
nn_predictions = data.frame(
  Actuals = test_prepped$loan_default)

# model class and probability predictions
nn1_class = predict(nn,newdata=test_prepped,type="raw")
nn_predictions$nn1_prob = predict(nn, newdata = test_prepped, type = "prob")[, "Yes"]

# computing F1
pred = prediction(nn_predictions$nn1_prob, test_prepped$loan_default)

prec = ROCR::performance(pred, "prec")
rec  = ROCR::performance(pred, "rec")

precision = prec@y.values[[1]]
recall    = rec@y.values[[1]]

# Compute F1 for each threshold
f1 = 2 * precision * recall / (precision + recall)
f1[is.nan(f1)] = 0


# combining F1 scores and cuttofs into df
cutoffs = prec@x.values[[1]]
df_f1 = data.frame(f1,cutoffs)
head(df_f1)

# extracting best f1
opt_idx = which.max(df_f1$f1)
opt_f1  = df_f1[opt_idx, ]
opt_f1 # best thresh is .2601289

best_thresh = opt_f1$cutoffs

# making best F1 prediction and adding to pred df
nn_predictions$nn1_class = ifelse(nn_predictions$nn1_prob >= best_thresh, "Yes", "No")
nn_predictions$nn1_class  = factor(nn_predictions$nn1_class, levels = c("Yes", "No"))



confusionMatrix(nn_predictions$nn1_class,test_prepped$loan_default ,positive="Yes")

```


## ROC Curve
```{r}

probabilities=predict(nn,newdata=test_prepped,type="prob")
probabilities$actual=test_prepped$loan_default

# Create Prediction Object 
pred = prediction(probabilities[,"Yes"], test_prepped$loan_default
                  )
# Create Performance Object
model1_perf = performance(pred, "tpr", "fpr")
auc = performance(pred,"auc")
auc_value=auc@y.values[[1]]

# Plot the ROC curve
plot(model1_perf, col = "blue", lwd = 2, main = "ROC Curve")
abline(a=0, b=1, col="red", lty=2)
legend("bottomright",legend=paste("AUC = ",round(auc_value,3)),col="blue",lwd=2)
```



## NN 2
```{r}
#cores = parallel::detectCores()
#cores
#cl = parallel::makeCluster(cores-1)
#registerDoParallel(cl)

#set.seed(123)

#tuneGrid = expand.grid(
#  size = c(12),
#  decay = c(.15)
#)


#trControl = trainControl(
#  method = "cv",
#  number = 5, # switch back to 5 after test 
#  classProbs = TRUE,
#  summaryFunction = twoClassSummary
#)
#nn2=train(loan_default ~ .,
#                 data = train_prepped,
#                  method = "nnet",
#                  trControl= trControl,
#                  metric="ROC",
#                  linout = FALSE,
#                  trace = FALSE,
#                  maxit = 300,
#                  tuneGrid = tuneGrid)
#nn2
#nn2$bestTune


#saveRDS(nn2, "nn_model2.rds")


#stopCluster(cl)
#registerDoSEQ()


```
First run (Size 5, 10, 15, Decay .1, .3, .5, .7) best Size = 10 and Decay = .3

Rerunning testing 8, 10, 12, and decays .2 - .5 by .05 ( previous best decay of .4, so testing in that range) 

Next best fit was 10 and .2 but best sensitivity occurred at 12 deay .3 -- Tuning again 10, 11, 12, 13 and decays .15 - .35 by .05

Fitting the best possible fit 



## Making Predictions

```{r}
nn2 = readRDS('nn_model2.rds')

# model class and probability predictions
nn_predictions$nn2_prob = predict(nn2, newdata = test_prepped, type = "prob")[, "Yes"]

# computing F1
model2_pred = prediction(nn_predictions$nn2_prob, test_prepped$loan_default)

prec = ROCR::performance(model2_pred, "prec")
rec  = ROCR::performance(model2_pred, "rec")

precision = prec@y.values[[1]]
recall    = rec@y.values[[1]]

# Compute F1 for each threshold
f1 = 2 * precision * recall / (precision + recall)
f1[is.nan(f1)] = 0


# combining F1 scores and cuttofs into df
cutoffs = prec@x.values[[1]]
df_f1 = data.frame(f1,cutoffs)
head(df_f1)

# extracting best f1
opt_idx = which.max(df_f1$f1)
opt_f1  = df_f1[opt_idx, ]
opt_f1 # best thresh is .2601289

best_thresh = opt_f1$cutoffs

# making best F1 prediction and adding to pred df
nn_predictions$nn2_class = ifelse(nn_predictions$nn2_prob >= best_thresh, "Yes", "No")
nn_predictions$nn2_class  = factor(nn_predictions$nn2_class, levels = c("Yes", "No"))



confusionMatrix(nn_predictions$nn2_class,test_prepped$loan_default ,positive="Yes")

```


## ROC Curve
```{r}

probabilities=predict(nn2,newdata=test_prepped,type="prob")
probabilities$actual=test_prepped$loan_default

# Create Prediction Object 
pred = prediction(probabilities[,"Yes"], test_prepped$loan_default
                  )
# Create Performance Object
model2_perf = performance(pred, "tpr", "fpr")
auc = performance(pred,"auc")
auc_value=auc@y.values[[1]]

# Plot the ROC curve
plot(model2_perf, col = "blue", lwd = 2, main = "ROC Curve")
abline(a=0, b=1, col="red", lty=2)
legend("bottomright",legend=paste("AUC = ",round(auc_value,3)),col="blue",lwd=2)
```
```{r}
#install.packages('keras')
```

## Attempting a Keras NN
### Using ChatGPT to help fit a Keras NN
```{r}
# =====================================================================
# FULL WORKING KERAS NEURAL NETWORK SCRIPT WITH TUNING LOOP (WITH DROPOUT)
# =====================================================================

library(keras)
library(pROC)

# -----------------------------
# 1. Prepare Data
# -----------------------------
x <- as.matrix(train_prepped[, -which(names(train_prepped) == "loan_default")])
y <- ifelse(train_prepped$loan_default == "Yes", 1, 0)

set.seed(123)

x_train <- as.matrix(train_prepped[, -which(names(train_prepped) == "loan_default")])
y_train <- ifelse(train_prepped$loan_default == "Yes", 1, 0)

x_test  <- as.matrix(test_prepped[, -which(names(test_prepped) == "loan_default")])
y_test  <- ifelse(test_prepped$loan_default == "Yes", 1, 0)


# -----------------------------
# 2. Hyperparameter Grid
# -----------------------------
units1_list <- c(16, 24, 32)
units2_list <- c(8, 12, 16, 24)
decay_list <- c(1e-05, 1e-04, 5e-04, 1e-03)
dropout_list <- c(0.1, 0.2, 0.3) # each node has a 90% chance to stay active, 80% chance, 70% chance

results <- data.frame(
  units1 = numeric(),
  units2 = numeric(),
  decay = numeric(),
  dropout = numeric(),
  auc = numeric(),
  stringsAsFactors = FALSE
)

# -----------------------------
# 3. Tuning Loop
# -----------------------------
for (u1 in units1_list) {
  for (u2 in units2_list) {
    for (d in decay_list) {
      for (dr in dropout_list) {

        cat("Testing: units1 =", u1, ", units2 =", u2, ", decay =", d, ", dropout =", dr, "\n")

        # Build model WITH DROPOUT
        model <- keras_model_sequential() %>%
          layer_dense(
            units = u1,
            activation = "relu",
            input_shape = ncol(x_train),
            kernel_regularizer = regularizer_l2(d)
          ) %>%
          layer_dropout(rate = dr) %>%                           # <---- ADDED DROP OUT

          layer_dense(
            units = u2,
            activation = "relu",
            kernel_regularizer = regularizer_l2(d)
          ) %>%
          layer_dropout(rate = dr) %>%                           # <---- ADDED DROP OUT

          layer_dense(units = 1, activation = "sigmoid")

        model %>% compile(
          optimizer = optimizer_adam(0.001),
          loss = "binary_crossentropy",
          metrics = "accuracy"
        )

        model %>% fit(
          x_train, y_train,
          epochs = 8,
          batch_size = 256,
          verbose = 0,
          validation_split = 0.2
        )

        pred_prob <- as.numeric(model %>% predict(x_test))
        auc_val <- auc(y_test, pred_prob)

        results <- rbind(
          results,
          data.frame(units1 = u1, units2 = u2, decay = d, dropout = dr, auc = auc_val)
        )
      }
    }
  }
}

# -----------------------------
# 4. Pick Best Model Based on AUC
# -----------------------------
best_row <- results[which.max(results$auc), ]
print(best_row)

best_u1 <- best_row$units1
best_u2 <- best_row$units2
best_decay <- best_row$decay
best_dropout <- best_row$dropout

cat("\nRetraining best model with units1 =", best_u1,
    "units2 =", best_u2,
    "decay =", best_decay,
    "dropout =", best_dropout, "\n")

# -----------------------------
# 5. Retrain Best Model Properly (WITH DROPOUT)
# -----------------------------
best_model <- keras_model_sequential() %>%
  layer_dense(
    units = best_u1,
    activation = "relu",
    input_shape = ncol(x_train),
    kernel_regularizer = regularizer_l2(best_decay)
  ) %>%
  layer_dropout(rate = best_dropout) %>%                         # <---- ADDED DROP OUT

  layer_dense(
    units = best_u2,
    activation = "relu",
    kernel_regularizer = regularizer_l2(best_decay)
  ) %>%
  layer_dropout(rate = best_dropout) %>%                         # <---- ADDED DROP OUT

  layer_dense(units = 1, activation = "sigmoid")

best_model %>% compile(
  optimizer = optimizer_adam(0.001),
  loss = "binary_crossentropy",
  metrics = c("accuracy", metric_auc())
)

history <- best_model %>% fit(
  x_train, y_train,
  epochs = 25,
  batch_size = 64,
  validation_split = 0.2,
  callbacks = callback_early_stopping(patience = 3, restore_best_weights = TRUE)
)

# -----------------------------
# 6. Evaluate Final Model
# -----------------------------
evaluation <- best_model %>% evaluate(x_test, y_test)
print(evaluation)

# -----------------------------
# 7. Predict Using Optimal F1 Threshold
# -----------------------------

nn_predictions$pred_prob <- as.numeric(best_model %>% predict(x_test))

thresholds <- seq(0, 1, by = 0.01)

f1_for_threshold <- function(t) {
  pred_temp <- ifelse(pred_prob > t, 1, 0)

  TP <- sum(pred_temp == 1 & y_test == 1)
  FP <- sum(pred_temp == 1 & y_test == 0)
  FN <- sum(pred_temp == 0 & y_test == 1)

  precision <- ifelse((TP + FP) == 0, 0, TP / (TP + FP))
  recall    <- ifelse((TP + FN) == 0, 0, TP / (TP + FN))

  if (precision + recall == 0) return(0)
  return(2 * precision * recall / (precision + recall))
}

f1_scores <- sapply(thresholds, f1_for_threshold)
best_threshold <- thresholds[which.max(f1_scores)]

cat("Best F1 threshold:", best_threshold, "\n")

nn_predictions$pred_class <- ifelse(pred_prob > best_threshold, 1, 0)

# Confusion Matrix
TP <- sum(nn_predictions$pred_class == 1 & y_test == 1)
FP <- sum(nn_predictions$pred_class == 1 & y_test == 0)
TN <- sum(nn_predictions$pred_class == 0 & y_test == 0)
FN <- sum(nn_predictions$pred_class == 0 & y_test == 1)

conf_matrix <- matrix(c(TN, FP, FN, TP), nrow = 2, byrow = TRUE)
colnames(conf_matrix) <- c("Pred 0", "Pred 1")
rownames(conf_matrix) <- c("Actual 0", "Actual 1")

cat("\nConfusion Matrix at Optimal F1 Threshold:\n")
print(conf_matrix)

# -----------------------------
# 8. Compute Final F1 Score
# -----------------------------
final_precision <- TP / (TP + FP)
final_recall    <- TP / (TP + FN)
final_f1        <- 2 * final_precision * final_recall / (final_precision + final_recall)

cat("\nFinal Precision:", round(final_precision, 4))
cat("\nFinal Recall:", round(final_recall, 4))
cat("\nFinal F1 Score:", round(final_f1, 4), "\n")



# =====================================================================
# END OF SCRIPT
# =====================================================================

saveRDS(best_model, "keras_nn.rds")

```

## ROC
```{r}


#install.packages("pROC")
library(pROC)

roc_obj <- roc(y_test, nn_predictions$pred_prob)
auc_value <- auc(roc_obj)

cat("AUC:", auc_value, "\n")

plot(roc_obj, col = "blue", lwd = 2, main = "ROC Curve")
abline(a = 1, b = -1, col = "red", lty = 2)

```

# NN 4

```{r}
cores = parallel::detectCores()
cores
cl = parallel::makeCluster(cores-1)
registerDoParallel(cl)

set.seed(123)

cv_control=trainControl(
  method="cv",
  number=5,
  classProbs = TRUE)

grid=expand.grid(size=c(1,2), decay=c(0.05,0.15))


nn4 <- train(
  loan_default ~ .,
  data=train_prepped,
  method="nnet",
  trControl = cv_control,
  linout = FALSE,
  trace = FALSE,
  tuneGrid=grid,
  maxit = 200,
  metric = "ROC"
)

nn4
nn4$bestTune


saveRDS(nn4, "nn_model4.rds")


stopCluster(cl)
registerDoSEQ()


```

## Making Predictions

```{r}
##
nn4 = readRDS('nn_model4.rds')


# model class and probability predictions
nn_predictions$nn4_prob = predict(nn4, newdata = test_prepped, type = "prob")[, "Yes"]

# computing F1
model4_pred = prediction(nn_predictions$nn4_prob, test_prepped$loan_default)

prec = ROCR::performance(model4_pred, "prec")
rec  = ROCR::performance(model4_pred, "rec")

precision = prec@y.values[[1]]
recall    = rec@y.values[[1]]

# Compute F1 for each threshold
f1 = 2 * precision * recall / (precision + recall)
f1[is.nan(f1)] = 0


# combining F1 scores and cuttofs into df
cutoffs = prec@x.values[[1]]
df_f1 = data.frame(f1,cutoffs)
head(df_f1)

# extracting best f1
opt_idx = which.max(df_f1$f1)
opt_f1  = df_f1[opt_idx, ]
opt_f1 

best_thresh = opt_f1$cutoffs

# making best F1 prediction and adding to pred df
nn_predictions$nn4_class = ifelse(nn_predictions$nn4_prob >= best_thresh, "Yes", "No")
nn_predictions$nn4_class  = factor(nn_predictions$nn4_class, levels = c("Yes", "No"))



confusionMatrix(nn_predictions$nn4_class,test_prepped$loan_default ,positive="Yes")

```

## ROC Curve
```{r}

probabilities=predict(nn4,newdata=test_prepped,type="prob")
probabilities$actual=test_prepped$loan_default

# Create Prediction Object 
pred = prediction(probabilities[,"Yes"], test_prepped$loan_default
                  )
# Create Performance Object
model4_perf = performance(pred, "tpr", "fpr")
auc = performance(pred,"auc")
auc_value=auc@y.values[[1]]

# Plot the ROC curve
plot(model4_perf, col = "blue", lwd = 2, main = "ROC Curve")
abline(a=0, b=1, col="red", lty=2)
legend("bottomright",legend=paste("AUC = ",round(auc_value,3)),col="blue",lwd=2)
```

# Comparing our models


## ROC Comparison
```{r}
plot(model1_perf ,col = "blue", lwd = 2 , main = "ROC Curves")
plot(model2_perf, col = "green", lwd = 2 , add = TRUE)
#plot(model3_perf, col = "orange", lwd = 2 , add = TRUE)
plot(model4_perf, col = "yellow", lwd = 2 , add = TRUE)



abline(a = 0, b = 1, col = "red", lty = 2, lwd = 2)

```

## Comparing key Metrics
```{r}
get_manual_metrics <- function(pred_class, pred_prob, actual, model_name) {
cm <- confusionMatrix(
factor(pred_class, levels = c("No", "Yes")),
factor(actual, levels = c("No", "Yes")),
positive = "Yes"
)

F1_val <- MLmetrics::F1_Score(y_pred = as.character(pred_class),
y_true = as.character(actual),
positive = "Yes")
AUC_val <- pROC::auc(ifelse(actual == "Yes", 1, 0),
as.numeric(pred_prob),
direction = "<")

data.frame(
Model = model_name,
Accuracy = round(cm$overall["Accuracy"], 4),
Recall = round(cm$byClass["Sensitivity"], 4),
Precision = round(cm$byClass["Pos Pred Value"], 4),
F1 = round(F1_val, 4),
Balanced_Accuracy = round(cm$byClass["Balanced Accuracy"], 4),
Detection_Rate = round(cm$byClass["Detection Rate"], 4),
AUC = round(as.numeric(AUC_val), 4)
)
}

metrics_model1 <- get_manual_metrics(nn_predictions$nn1_class, nn_predictions$nn1_prob, nn_predictions$Actuals, "model1")
metrics_model2  <- get_manual_metrics(nn_predictions$nn2_class,   nn_predictions$nn2_prob,   nn_predictions$Actuals, "model2")
metrics_model3  <- get_manual_metrics(nn_predictions$pred_class,  nn_predictions$pred_prob, nn_predictions$Actuals, "model3")
metrics_model4  <- get_manual_metrics(nn_predictions$nn4_class,  nn_predictions$nn4_prob, nn_predictions$Actuals, "model4")

model_comparison <- rbind(metrics_model1, metrics_model2, metrics_model3, metrics_model4) 
model_comparison
```
Keras Model Results:
Final Precision: 0.4524
Final Recall: 0.614
Final F1 Score: 0.521 

Our best performing model was model2, with the highest recall (67%), highest F1 score (.556) and highest AUC (.8252). Model 2 outperforms all other models given our task of correctly identifying loan defaults. All of these models appear to perform similarly to our regularized regression models.

